---
title: "Reproducible Research: Peer Assessment 1"
Author: Liana Engie
output: 
  html_document:
    keep_md: true
---

#Assumes in your working directory

## Loading and preprocessing the data

First, load the requisite packages in R. 
Then, download the data from the internet and unzip it.

```{r}
library(dplyr)
library(knitr)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","act.zip")
unzip("act.zip")
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

To get a sense of the data, the sum of total steps per day was calculated, as can be seen in the following graph. 
A histogram was created of the frequency of number of steps taken in a day.

```{r}
by_date <- group_by(data,date)
total_steps <- summarize(by_date,total=sum(steps,na.rm=TRUE))
#print(total_steps)

s <- ggplot(total_steps,aes(date,total,group=1))+
     geom_point(size=3)+
     geom_line(,size=.5)+
     theme_gray()+
     theme(axis.text.x=element_text(angle=80))+
     scale_x_discrete(
          c("2012-10-01",
            "2012-10-15",
            "2012-10-31",
            "2012-11-15",
            "2012-11-30"))
print(s)

h <- ggplot(total_steps,aes(total))+
     geom_histogram(color="steelblue")+
     theme_gray()+
     ylab("Frequency")+
     xlab("Total steps per day")+
     ggtitle("Steps per Day in October-Nov 2012")
print(h)
```

From this data, the mean and median number of steps per day can be calculated.

```{r}
overall_mean <- mean(total_steps$total)
print(overall_mean)

median <- median(total_steps$total)
print(median)
```

## What is the average daily activity pattern?
```{r}
by_interval <- data %>% group_by(interval)
by_interval %>% summarize(total=sum(steps,na.rm=TRUE))
print(by_interval)
```


## Inputing missing values



## Are there differences in activity patterns between weekdays and weekends?
